---
title: "Introduction to dtasamplesize: Beyond Classical Sample Size Formulas for Diagnostic Accuracy Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dtasamplesize: Beyond Classical Sample Size Formulas for Diagnostic Accuracy Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why classical formulas are insufficient

The Buderer (1996) formula is the most widely used method for computing
sample sizes in diagnostic accuracy studies. However, Monte Carlo
simulation reveals it achieves the target precision only about 57% of
the time:

```{r buderer-validation}
library(dtasamplesize)
result <- mc_validate_buderer(Se = 0.85, d = 0.07, B = 500, seed = 2026)
result$results[, c("ci_method", "P_width_target", "coverage")]
```

With Se = 0.85 and desired half-width d = 0.07, Buderer yields
n = `r result$n_diseased` diseased subjects. Yet the probability that
the CI width actually falls within the target is only ~57%, far below
the 95% one might expect.

## Bayesian Assurance Method

The BAM (Wilson et al. 2022) incorporates prior information and searches
for the sample size achieving a specified assurance probability:

The search grid `n_range` must be wide enough to actually contain the answer
for **both** arms. With the vague default `prior_sp = c(2, 2)` and
`delta_sp = 0.10`, the specificity arm needs about 370 non-diseased subjects,
so a grid that stops at 200 would silently fall back to its own upper limit
(with a warning) and understate the total N.

```{r bam}
bam_result <- bam_sample_size(
  prior_se = c(17, 3), delta_se = 0.14,
  B = 500, seed = 2026, n_range = 20:600
)
cat("BAM n_diseased:", bam_result$n_diseased, "\n")
cat("BAM n_non_diseased:", bam_result$n_non_diseased, "\n")
cat("Assurance Se:", round(bam_result$assurance_se, 3), "\n")
```

## Joint optimization of Se, Sp, and AUC

In practice, investigators need adequate precision for sensitivity,
specificity, and AUC simultaneously. The joint approach finds the
minimum N satisfying all three constraints:

Note that `Se`, `Sp` and `AUC` are not free parameters: any concave ROC
curve through the operating point `(1 - Sp, Se)` has area at least
`0.5*(1-Sp)*Se + 0.5*Sp*(Se+1)`, which is 0.875 at the defaults
`Se = 0.85` and `Sp = 0.90`. `joint_sample_size()` refuses an `AUC` below
that bound rather than sizing a study for a test that cannot exist.

We fix the prevalence at 0.30 here so that every method in the comparison
below is evaluated at the *same* prevalence.

```{r joint}
joint_result <- joint_sample_size(
  B = 500, seed = 2026, prev = 0.30, N_range = seq(100, 700, by = 50)
)
cat("Joint N:", joint_result$n_total,
    "vs Buderer N:", joint_result$buderer_N, "\n")
```

## Imperfect reference standard

When the reference standard is imperfect, the Rogan-Gladen variance
inflation factor (VIF) inflates the required sample size:

```{r imperfect}
imp_result <- ss_imperfect_ref(B = 0)
imp_result$sensitivity_table[, c("Se_ref", "Sp_ref", "VIF", "N_adj")]
```

## Comparison across methods

All four approaches below are evaluated at the **same disease prevalence of
0.30**: the classical and joint calculations take `prev = 0.30` directly, and
the Bayesian assurance method uses its default `Beta(6, 14)` prevalence prior,
whose mean is 0.30. Comparing methods that were sized at *different*
prevalences would confound the method with the prevalence, since the required
total N scales roughly as `1 / prev` on the sensitivity arm.

```{r comparison}
comparison <- data.frame(
  Method = c("Buderer (classical)", "BAM (Bayesian)",
             "Joint (Se+Sp+AUC)", "Imperfect ref (Rogan-Gladen)"),
  N = c(
    ceiling(buderer_n(0.85, 0.07) / 0.30),
    bam_result$N_total_median,
    joint_result$n_total,
    imp_result$N_adjusted_loss
  )
)
comparison
```

The unified framework (`ss_unified()`) integrates all sources of
uncertainty into a single simulation. See `?ss_unified` for details.

## Sensitivity analysis across prior scenarios

The prior beliefs about Se and Sp that go into `ss_unified()` are rarely known
with certainty at the design stage. `sensitivity_analysis()` runs
`ss_unified()` once per named scenario and collects the results into a single
table, replacing a hand-written loop over scenarios. Below, an optimistic, a
neutral, and a pessimistic scenario vary only in how confident (and how high)
the assumed Se and Sp are; the disease-prevalence prior is held fixed so that
the comparison isolates the effect of the accuracy assumptions.

```{r sensitivity}
scenarios <- list(
  optimistic = list(prior_se = c(18, 2), prior_sp = c(19, 1),
                     prior_prev = c(5, 15)),
  neutral = list(prior_se = c(17, 3), prior_sp = c(2, 2),
                  prior_prev = c(5, 15)),
  pessimistic = list(prior_se = c(14, 6), prior_sp = c(16, 4),
                      prior_prev = c(5, 15))
)

# B and N_range are kept small here for a fast vignette build; see
# ?sensitivity_analysis and the 'B' argument of ss_unified() for guidance
# on the B recommended for a final analysis.
sens_result <- suppressWarnings(sensitivity_analysis(
  scenarios,
  delta_se = 0.07, delta_sp = 0.05, delta_auc = 0,
  N_range = seq(300, 1200, by = 100), B = 300, seed = 2026
))
sens_result
```

Reading the table: `N_total` is the loss-adjusted total sample size the
unified framework recommends for that scenario, and `joint_assurance` is the
probability actually achieved at that N (at or above the 0.80 target).
`N_buderer` is the classical Buderer total for the same scenario, included so
the gap between the assurance-based and classical requirements can be seen
at a glance for each set of assumptions.

The pattern is the one intuition predicts: the pessimistic scenario, which
assumes lower and less certain Se and Sp, needs the largest sample size,
while the optimistic scenario needs the smallest. What the table adds is the
*size* of that gap -- here several hundred subjects between the two extremes
-- which is exactly the information a protocol needs when the "expected"
accuracy is a guess rather than a certainty: it shows how much the required
sample size would grow if the study turns out to be in the harder-to-detect
regime, rather than reporting a single N that is silent about that risk.
