---
title: "Introduction to dtasamplesize: Beyond Classical Sample Size Formulas for Diagnostic Accuracy Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dtasamplesize}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why classical formulas are insufficient

The Buderer (1996) formula is the most widely used method for computing
sample sizes in diagnostic accuracy studies. However, Monte Carlo
simulation reveals it achieves the target precision only about 57% of
the time:

```{r buderer-validation}
library(dtasamplesize)
result <- mc_validate_buderer(Se = 0.85, d = 0.07, B = 500, seed = 2026)
result$results[, c("ci_method", "P_width_target", "coverage")]
```

With Se = 0.85 and desired half-width d = 0.07, Buderer yields
n = `r result$n_diseased` diseased subjects. Yet the probability that
the CI width actually falls within the target is only ~57%, far below
the 95% one might expect.

## Bayesian Assurance Method

The BAM (Wilson et al. 2022) incorporates prior information and searches
for the sample size achieving a specified assurance probability:

```{r bam}
bam_result <- bam_sample_size(
  prior_se = c(17, 3), delta_se = 0.14,
  B = 500, seed = 2026, n_range = 20:200
)
cat("BAM n_diseased:", bam_result$n_diseased, "\n")
cat("Assurance Se:", round(bam_result$assurance_se, 3), "\n")
```

## Joint optimization of Se, Sp, and AUC

In practice, investigators need adequate precision for sensitivity,
specificity, and AUC simultaneously. The joint approach finds the
minimum N satisfying all three constraints:

```{r joint}
joint_result <- joint_sample_size(
  B = 500, seed = 2026, N_range = seq(100, 700, by = 50)
)
cat("Joint N:", joint_result$n_total,
    "vs Buderer N:", joint_result$buderer_N, "\n")
```

## Imperfect reference standard

When the reference standard is imperfect, the Staquet correction inflates
the required sample size via a variance inflation factor (VIF):

```{r imperfect}
imp_result <- ss_imperfect_ref(B = 0)
imp_result$sensitivity_table[, c("Se_ref", "Sp_ref", "VIF", "N_adj")]
```

## Comparison across methods

```{r comparison}
comparison <- data.frame(
  Method = c("Buderer (classical)", "BAM (Bayesian)",
             "Joint (Se+Sp+AUC)", "Imperfect ref"),
  N = c(
    ceiling(buderer_n(0.85, 0.07) / 0.30),
    bam_result$N_total_median,
    joint_result$n_total,
    imp_result$N_adjusted_loss
  )
)
comparison
```

The unified framework (`ss_unified()`) integrates all sources of
uncertainty into a single simulation. See `?ss_unified` for details.
